---
title: "ms1Rscript_NOAA"
output: pdf_document
date: '2022-07-29'
---
# This file contains R code for establishing the priors used in the BHM2 and BHM3 models. The priors are established using the first 9 tests.
```{r}
packages<-function(x, repos="http://cran.r-project.org", ...){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x, repos=repos, ...)
    require(x,character.only=TRUE)
  }
}

base <- getwd()

packages(rv)
packages(rstan)
packages(tidyverse)
packages(tikzDevice)
packages(reshape())


rstan_options(auto_write = TRUE)
options(mc.cores = min(c(parallel::detectCores(), 8)))

nchains <-  min(c(parallel::detectCores(), 8))
niters <- 50000
nkeep <- 2500
nthin <- ceiling((niters/2)*nchains/nkeep)
```

Load NOAA Data - the code for organizing the data can be found in the file called "Reading_Data_Sheets.Rmd"
```{r}
# file names
load("NOAA_data_names.RData")

# load linear data sheets
load("all_NOAA_linear.RData")

# load nonlinear data sheets
load("all_NOAA_nonlinear.RData")

## df for all linear templates 
NOAA1<-merge_recurse(lineardata[1:83])

## df for all nonlinear templates 
NOAA2<-merge_recurse(nonlineardata[1:131])

## df for all nonlinear templates 
NOAA<-rbind(NOAA1, NOAA2)

#Only keep first nine tests for establishing initial priors
NOAA<-subset(NOAA, NOAA$Test<10)

# change columns names so the sample IDs w/ dilution factors are in column "SampleID"
names(NOAA)<-c("SampleID2", "Absorbance", "Concentration", "Dilution", "Test","NOAAconc", "SampleID")

```
Code for 'controlS'
```{r}
stan.in3 <- function(data = NOAA, chains=nchains){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    y <- data$Absorbance[!temp]
    x <- data$Concentration[!temp]
    N <- n-sum(temp)
    M <- sum(temp)
    y0 <- data$Absorbance[temp]
    dlf <- data$Dilution[temp]
    wsC<-ordered(data$SampleID[temp])
    ws <- as.numeric(wsC)
    stan.dat <- list(N=N, M=M, MM=max(ws), y=y, x=x,
                     y0=y0, ws=ws, dlf=dlf)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(delta = runif(1,0.5,1), th2=runif(1),
                           th3 = runif(1), th1=runif(1,0,0.1),
                           logx0 = rnorm(max(ws)),
                           sigma = runif(1))
    parameters <- c("th1","th2","th3","th4","sigma","x0")
    return(list(para=parameters, data=stan.dat, inits=inits,
              n.chains=chains, ControlSample=(1:max(ws))[levels(wsC)=="Control 1"]))
}


controlS <- numeric()
for (i in 1:9){
    print(paste("Test:", i))
    input.to.stan <- stan.in3(data=NOAA[NOAA$Test==i,])
    controlS[i] <- input.to.stan$ControlSample
}
```

Only do 9 tests. Takes very long to run all tests. sBu avoids this long process. This model provides initial priors for sBu models BHM2 and BHM3.
```{r}
### Hierarchical ELISA model with water samples
###    unknown concentrations estimaterd using hierarchical model
### Stan Model (multiple test kits)
### The model: y=th4+(th1-th4)/[1+(x/th3)^(-th2)]

ELISA4 <- "
data {
  int<lower=0> N; // sample size
  int<lower=0> K; // number of test kits
  int<lower=0> M; //observed ODs (for estimating concentration)
  real y[N]; // std solution response data
  real x[N]; // std solution conc
  real y0[M]; // test data response
  int MM; //unique samples
  int ws[M]; //unique water samples
  int<lower=1,upper=K> test1[N];
  int<lower=1,upper=K> test2[M];
  int<lower=1,upper=K> test3[MM];
}
parameters {
  vector[K] zdelta;
  vector[K] zth1;
  vector[K] zth2;
  vector[K] zth3;
  vector<lower=0>[K] sigma;
  vector[MM] logx0;
  vector<lower=0>[K] x0sig;
  vector[K] x0mu;
  real mu_dlt;
  real<lower=0> sig_dlt;
  real mu_th1;
  real<lower=0> sig_th1;
  real mu_th2;
  real<lower=0> sig_th2;
  real mu_th3;
  real<lower=0> sig_th3;
}
transformed parameters{
  vector<lower=0>[K] delta;
  vector<lower=0>[K] th1;
  vector<lower=0>[K] th2;
  vector<lower=0>[K] th3;
  vector[N] mu;
  vector[M] mu0;
  vector<lower=0>[N] sig;
  vector<lower=0>[M] sig0;
  vector<lower=0>[K] th4;
  vector<lower=0>[MM] x0;

  x0 = exp(logx0);
  delta = mu_dlt+sig_dlt*zdelta;
  th1 = mu_th1+sig_th1*zth1;
  th2 = mu_th2+sig_th2*zth2;
  th3 = mu_th3+sig_th3*zth3;
  th4 = th1+delta;
  for (i in 1:N){
    mu[i] = th4[test1[i]] -
              delta[test1[i]]/(1+(x[i]/th3[test1[i]])^(-th2[test1[i]]));
    sig[i] = sigma[test1[i]];
  }
  for (i in 1:M){
    mu0[i] = th4[test2[i]] -
        delta[test2[i]]/(1+(x0[ws[i]]/th3[test2[i]])^(-th2[test2[i]]));
    sig0[i] = sigma[test2[i]];
  }
}
model {
  x0mu ~ normal(0, 2.5);
  x0sig ~ normal(0, 2.5);
  mu_dlt ~ normal(0, 2.5);
  sig_dlt ~ normal(0, 2.5);
  mu_th1 ~ normal(0, 2.5);
  sig_th1 ~ normal(0, 2.5);
  mu_th2 ~ normal(0, 2.5);
  sig_th2 ~ normal(0, 2.5);
  mu_th3 ~ normal(0, 2.5);
  sig_th3 ~ normal(0, 2.5);

  for (j in 1:MM)
    logx0[j] ~ normal(x0mu[test3[j]], x0sig[test3[j]]);
  zth1~std_normal();
  zth2~std_normal();
  zth3~std_normal();
  zdelta~std_normal();
  target += normal_lpdf(y| mu, sig);
  target += normal_lpdf(y0| mu0, sig0);
}
"

stan.fit4 <- stan_model(model_code=ELISA4)

stan.in5 <- function(data = NOAA, chains=nchains){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    tmp1 <- data[!temp,]
    tmp2 <- data[temp,]
    test1 <- as.numeric(ordered(tmp1$Test))
    y <- tmp1$Absorbance
    x <- tmp1$Concentration
    N <- dim(tmp1)[1]
    M <- dim(tmp2)[1]
    K <- max(test1)
    y0 <- tmp2$Absorbance
    wsC <- ordered(paste(tmp2$Test,tmp2$SampleID2))
    wsLevels <- levels(wsC)
    tmpp <- str_detect(wsLevels, "Control")
    ws <- as.numeric(wsC)
    test2 <- as.numeric(ordered(tmp2$Test))
    test3 <- as.numeric(substring(wsLevels, 1, 1))
    stan.dat <- list(N=N, M=M, MM=max(ws), K=K, y=y, x=x,
                     y0=y0, ws=ws, test1=test1,
                     test2=test2, test3=test3)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(zdelta=runif(K,0.5,1), zth1=runif(K,0,0.1),
                           zth2=runif(K,0.4,0.6), zth3=runif(K,0.3,0.7),
                           sigma=runif(K), logx0=rnorm(max(ws)),
                           x0sig=runif(K), x0mu=rnorm(K),
                           mu_dlt=runif(1), sig_dlt=runif(1),
                           mu_th1=runif(1), sig_th1=runif(1),
                           mu_th2=runif(1), sig_th2=runif(1),
                           mu_th3=runif(1), sig_th3=runif(1))

    parameters <- c("th1", "th2","th3","th4","sigma","x0",
                    "x0mu","x0sig", "mu_dlt", "sig_dlt", "mu_th1",
                    "sig_th1", "mu_th2", "sig_th2", "mu_th3", "sig_th3")
    return(list(para=parameters, data=stan.dat, inits=inits, n.chains=chains,
                ControlSample=(1:max(ws))[tmpp]))
}

input.to.stan <- stan.in5(data=NOAA)
BHM2ControlSamples <- input.to.stan$ControlSample
fit2keep <- sampling(stan.fit4, data = input.to.stan$data,
                     init=input.to.stan$inits,
                     pars = input.to.stan$para,
                     iter=niters, thin=nthin,
                     chains=input.to.stan$n.chains)
print(fit2keep)
test_stanBHM2 <- rstan::extract(fit2keep)
save(test_stanBHM2, file="BHM_Prior.RData")
load("BHM_Prior.RData")
```
