---
title: "ELISA_manuscript_plots"
author: "Sabrina"
date: '2023-10-03'
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r}
packages<-function(x, repos="http://cran.r-project.org", ...){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x, repos=repos, ...)
    require(x,character.only=TRUE)
  }
}

base <- getwd()

packages(tidyverse)
packages(tikzDevice)
packages(reshape())
packages(rv)
packages(rstan)

nchains <-  min(c(parallel::detectCores(), 8))
niters <- 50000
nkeep <- 2500
nthin <- ceiling((niters/2)*nchains/nkeep)

```

This file contains the code to create the two plots using tikzDevice for our ELISA manuscript.

# Standard sample plot

#Load NOAA Data
```{r}
# file names
load("NOAA_data_names.RData")

# load linear data sheets
load("all_NOAA_linear.RData")

# load nonlinear data sheets
load("all_NOAA_nonlinear.RData")

## df for all linear templates 
NOAA1<-merge_recurse(lineardata[1:83])

## df for all nonlinear templates 
NOAA2<-merge_recurse(nonlineardata[1:131])

## df for all nonlinear templates 
NOAA<-rbind(NOAA1, NOAA2)

# change columns names so the sample IDs w/ dilution factors are in column "SampleID"
names(NOAA)<-c("SampleID2", "Absorbance", "Concentration", "Dilution", "Test", "NOAAconc", "SampleID")

```


```{r}
# keep only standard samples
SS<-NOAA[!is.na(NOAA$Concentration),]
SS$number<-1:nrow(SS)

SS1<-subset(SS, SampleID2 %in% 'Standard 1') # locate 3 standard 1 samples that are the highest, lowest, and middle absorbances for the 0 concentration sample
#median(SS1$Absorbance)

SS2<-subset(SS1, number %in% c('775', '1094', '2342')) # Selecting the tests of the largest, middle, and smallest value standard 1 samples

SS3<-subset(SS, Test %in% c('65', '92', '196')) # subset all of the standards from these tests

tmp1<-subset(SS, Test %in% '65')
tmp1$id<-1
tmp2<-subset(SS, Test %in% '92')
tmp2$id<-2
tmp3<-subset(SS, Test %in% '196')
tmp3$id<-3

SS3<-rbind(tmp1, tmp2, tmp3)

SS4<-SS[!(SS$Test=="65" | SS$Test=="92"| SS$Test=="196"),]
SS4$id<-0

SS5<-rbind(SS4, SS3)

##tiff("standard_samples.tiff", units="in", width=3, height=3, res=300)
ggplot(SS5, aes(x=Concentration, y=Absorbance, group=as.factor(Test))) +
   geom_jitter(
    data = SS5[SS5$id == 0, ],
    col="grey") + geom_point(data = SS5 %>% filter(id %in% '1'), shape=15, size=2)  + geom_point(data = SS5 %>% filter(id %in% '2'), shape=17, size=2) + geom_point(data = SS5 %>% filter(id %in% '3'), shape=18, size=2) + stat_summary(data = SS5 %>% filter(id %in% '1'), geom = "line", fun = mean)+stat_summary(data = SS5 %>% filter(id %in% '2'), geom = "line", fun = mean)+ stat_summary(data = SS5 %>% filter(id %in% '3'), geom = "line", fun = mean)+ theme(legend.position = "none") + scale_color_manual(values = c( "black", "black", "black")) + xlab("Concentration (μg/L)")
##dev.off()
```
Tikz plot
```{r}
tikz(file="standardcurve_plot_jitter.tex", # file = name you want to save plot as
     height=2, width=3, standAlone=T)
ggplot(SS5, aes(x=Concentration, y=Absorbance, group=as.factor(Test))) +
   geom_jitter(
    data = SS5[SS5$id == 0, ],
    col="grey") + geom_point(data = SS5 %>% filter(id %in% '1'), shape=15, size=2)  + geom_point(data = SS5 %>% filter(id %in% '2'), shape=17, size=2) + geom_point(data = SS5 %>% filter(id %in% '3'), shape=18, size=2) + stat_summary(data = SS5 %>% filter(id %in% '1'), geom = "line", fun = mean)+stat_summary(data = SS5 %>% filter(id %in% '2'), geom = "line", fun = mean)+ stat_summary(data = SS5 %>% filter(id %in% '3'), geom = "line", fun = mean)+ theme(legend.position = "none") + scale_color_manual(values = c( "black", "black", "black"))  + xlab("Concentration ($\\mu$g/L)")
dev.off()
```

# 6 model comparison plot
Load results
```{r}
# model 1
load("BC5.RData") 
load("pred_summBC5.RData")
load("pred_rvsBC5.RData")
# model 2
load("BC12.RData") 
load("pred_summBC12.RData")
load("pred_rvsBC12.RData")
# model 3
load("Bayes.RData")
# model 4
load("BHM1.RData")
# model 5
load("BHM2.RData")
# model 6
load("BHM3.RData")
```
load in 'controlS' which are needed for the plots. This code comes from the Bayes model.
```{r}
stan.in3 <- function(data = NOAA, chains=nchains){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    y <- data$Absorbance[!temp]
    x <- data$Concentration[!temp]
    N <- n-sum(temp)
    M <- sum(temp)
    y0 <- data$Absorbance[temp]
    wsC<-ordered(data$SampleID[temp])
    ws <- as.numeric(wsC)
    stan.dat <- list(N=N, M=M, MM=max(ws), y=y, x=x,
                     y0=y0, ws=ws)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(delta = runif(1,0.5,1), th2=runif(1),
                           th3 = runif(1), th1=runif(1,0,0.1),
                           logx0 = rnorm(max(ws)),
                           sigma = runif(1))
    parameters <- c("th1","th2","th3","th4","sigma","x0")
    return(list(para=parameters, data=stan.dat, inits=inits,
              n.chains=chains, ControlSample=(1:max(ws))[levels(wsC)=="Control 1"]))
}

controlS <- numeric()
for (i in 1:214){
    #print(paste("Test:", i))
    input.to.stan <- stan.in3(data=NOAA[NOAA$Test==i,])
    if(length(input.to.stan$ControlSample)==0) # added 8/2/22
    controlS[i]<-0
    else
    controlS[i] <- input.to.stan$ControlSample
}
```
Compare the control estimates for each model bc controls have a known concentration of 0.75.
```{r}
## comparing control sample fits

cntrlID <- (1:length(pred_summ12[[1]]))[substring(pred_summ12[[i]]$SampleID, 1, 9)=="Control 1"]
CntrlMLE12 <- pred_summ12[[1]][cntrlID,]
CntrlMLErvs12 <- pred_rvs12[[1]][cntrlID]

for (i in 2:214){
    cntrlID <- (1:length(pred_summ12[[i]]))[substring(pred_summ12[[i]]$SampleID, 1, 9)=="Control 1"]
    CntrlMLE12 <- rbind(CntrlMLE12[,1:9], pred_summ12[[i]][cntrlID,1:9])
    CntrlMLErvs12 <- c(CntrlMLErvs12, pred_rvs12[[i]][cntrlID])
}

cntrlID <- (1:dim(pred_summ5[[1]])[1])[substring(pred_summ5[[1]]$SampleID, 1, 9)=="Control 1"]
CntrlMLE5 <- pred_summ5[[1]][cntrlID,]
CntrlMLErvs5 <- pred_rvs5[[1]][cntrlID]
for (i in 2:214){
    cntrlID <- (1:dim(pred_summ5[[i]])[1])[substring(pred_summ5[[i]]$SampleID, 1, 9)=="Control 1"]
    CntrlMLE5 <- rbind(CntrlMLE5[,1:9], pred_summ5[[i]][cntrlID,1:9])
    CntrlMLErvs5 <- c(CntrlMLErvs5, pred_rvs5[[i]][cntrlID])
}

##Bayesian models
# MLE
CntrlStan1 <- rvsims(test_stanMLE1[[1]]$x0)[controlS[1]]
for (i in 2:214)
CntrlStan1 <- c(CntrlStan1, rvsims(test_stanMLE1[[i]]$x0)[controlS[i]])

# within-test only
CntrlStan3 <- rvsims(test_stanBHM1[[1]]$x0)[controlS[1]]
for (i in 2:214)
CntrlStan3 <- c(CntrlStan3, rvsims(test_stanBHM1[[i]]$x0)[controlS[i]])

# across test only sequential BHM (starts from test 10)
CntrlStanSeq2 <- rvsims(seqBHM2[[10]]$x0)[controlS[10]]
for (i in 11:214)
CntrlStanSeq2 <- c(CntrlStanSeq2, rvsims(seqBHM2[[i]]$x0)[controlS[i]])

# within and across test sequential BHM (starts from test 10)
CntrlStanSeq3 <- rvsims(seqBHM3[[10]]$x0)[controlS[10]]
for (i in 11:214)
CntrlStanSeq3 <- c(CntrlStanSeq3, rvsims(seqBHM3[[i]]$x0)[controlS[i]])

```
Evaluating biasness/accuracy (0.75)
```{r}
### using median ###

accuracy <- c(median(abs(CntrlStanSeq3-0.75)),
          median(abs(CntrlStanSeq2-0.75)),
          median(abs(CntrlStan3-0.75)),
          median(abs(CntrlStan1-0.75)),
          median(abs(CntrlMLErvs12-0.75),na.rm=T),
          median(abs(CntrlMLErvs5-0.75),na.rm=T))
names(accuracy) <- c("$BHM_3$","$BHM_2$" ,"$BHM_1$",
                 "Bayes","$IFE_{12}$", "$IFE_5$")

#tiff("6_models.tiff", units="in", width=4.75, height=3.5, res=300)
par(mar=c(3, 3, 2, 0.125), mgp=c(1.25,0.25,0), tck=-0.01)
mlplot(accuracy, xlab="Accuracy",  xlim = c(0.08, 0.35), main="")
#dev.off
```

Tikz plot
```{r}
tikz(file="6_model_compare.tex", # file = name you want to save plot as
     height=3, width=3, standAlone=T)
par(mar=c(3, 3, 2, 1.25), mgp=c(1.25,0.25,0), tck=-0.01)
mlplot(accuracy, xlab="Accuracy", xlim=c(0.08, 0.35), main="")
dev.off()
```

