---
title: "ELISA_manuscript"
output: html_document
date: "2023-04-21"
---

# This file contains all of the r code for the different models used to evaluate NOAA-GLERL's ELISA tests. Code for comparison plots is also included.

```{r, include=FALSE}
packages<-function(x, repos="http://cran.r-project.org", ...){
  x<-as.character(match.call()[[2]])
  if (!require(x,character.only=TRUE)){
    install.packages(pkgs=x, repos=repos, ...)
    require(x,character.only=TRUE)
  }
}

base <- getwd()


packages(tidyverse)
packages(tikzDevice)
packages(reshape())
packages(rv)
packages(rstan)

rstan_options(auto_write = TRUE)
options(mc.cores = min(c(parallel::detectCores(), 8)))

nchains <-  min(c(parallel::detectCores(), 8))
niters <- 50000
nkeep <- 2500
nthin <- ceiling((niters/2)*nchains/nkeep)
```

#Load NOAA Data
```{r}
# file names
load("NOAA_data_names.RData")

# load linear data sheets
load("all_NOAA_linear.RData")

# load nonlinear data sheets
load("all_NOAA_nonlinear.RData")

## df for all linear templates 
NOAA1<-merge_recurse(lineardata[1:83])

## df for all nonlinear templates 
NOAA2<-merge_recurse(nonlineardata[1:131])

## df for all nonlinear templates 
NOAA<-rbind(NOAA1, NOAA2)

# change columns names so the sample IDs w/ dilution factors are in column "SampleID"
names(NOAA)<-c("SampleID2", "Absorbance", "Concentration", "Dilution", "Test", "NOAAconc", "SampleID")
```

#Models
7 different models:
1) back-calculated 6 data points (IFE 5)
2) back-calculated 12 data points (IFE 12)
3) Bayesian (non-informative prior; Bayes)
4) BHM within test (BHM 1)
5) BHM across test (sequential updating; BHM 2) 
6) BHM within and across test (sequential updating; BHM 3)

# 2) back-calculated 12 data points (IFE 12)
Bayesian fitted curve with no non-informative prior - this allows us to quanitfy a realistic uncertainty estimate for all of the coefficients. This still compares to their actual fitting process.
```{r}
##########Part 1: Stan models and rstan implementation ###########
### Default ELISA model -- simulating nonlinear regression
###  with standard solutions only
### Stan Model (single test kit)
### The model: y=th4+(th1-th4)/[1+(x/th3)^(-th2)]
ELISA1 <- "
data {
  int<lower=0> N; // sample size (training data, non-zero)
  real y[N]; // response data
  real x[N]; // predictor data
}
parameters {
  real<lower=0> delta;
  real<lower=0> th1;
  real<lower=0> th2;
  real<lower=0> th3;
  real<lower=0> sigma;
}
transformed parameters{
  real mu[N];
  real<lower=0> th4;
  th4 = th1+delta;
  for (i in 1:N){
    mu[i] = th4 - delta/(1+(x[i]/th3)^(-th2));
  }
}
model {
  delta~normal(1,5);
  th1~normal(0,5);
  th2~normal(0,5);
  th3~normal(0,5);
  target += normal_lpdf(y| mu, sigma);
}
"
stan.fit1 <- stan_model(model_code=ELISA1)

stan.in1 <- function(data = NOAA, chains=nchains){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    y <- data$Absorbance[!temp]
    x <- data$Concentration[!temp]
    N <- n-sum(temp)
    stan.dat <- list(N=N, y=y, x=x)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(delta = runif(1,0.5,1), th2=runif(1),
                           th3 = runif(1), th1=runif(1,0,0.2),
                           sigma = runif(1))
    parameters <- c("th1", "th2","th3","th4","sigma")
    return(list(para=parameters, data=stan.dat, inits=inits,
                n.chains=chains))
}

test_stan1 <- list()
for (i in 1:214){
    print(paste("Test:", i))
    input.to.stan <- stan.in1(data=NOAA[NOAA$Test==i,])
    fit2keep <- sampling(stan.fit1, data = input.to.stan$data,
                         init=input.to.stan$inits,
                         pars = input.to.stan$para,
                         iter=niters, thin=nthin,
                         chains=input.to.stan$n.chains)
    print(fit2keep)
    test_stan1[[i]] <- rstan::extract(fit2keep)
}
save(test_stan1, file="BC12.RData")
load("BC12.RData")

plot(test_stan1[[1]]$th1, log(test_stan1[[1]]$sigma))
## Neal's funnel -- explaining the reason of high levels of
##  uncertainty when using small sample sizes
##  Add a case where only five data points were used
## use the ELISA test data with all known concentrations to
##  show if the Neal's funnel disappears
```
# 1) back-calculated 5 data points (IFE 5)
```{r}
## five data point case

NOAA5 <- NOAA[!is.na(NOAA$Concentration), ]
NOAA5mean <- data.frame(
    Absorbance=tapply(NOAA5$Absorbance,
                      paste(NOAA5$SampleID, NOAA5$Test),
                      mean),
    Concentration=tapply(NOAA5$Concentration,
                      paste(NOAA5$SampleID, NOAA5$Test),
                      mean),
    Test=tapply(NOAA5$Test,
                      paste(NOAA5$SampleID, NOAA5$Test),
                      mean))

stan.in2 <- function(data = NOAA5mean, chains=nchains){
    Abs0 <- data$Absorbance[data$Concentration==0] 
    data <- data[data$Concentration!=0,]
    n <- dim(data)[1]
    y <- data$Absorbance/Abs0 # divide each non-zero standard solution avg abs by the zero concentration standard solution avg abs
    x <- data$Concentration
    stan.dat <- list(N=n, y=y, x=x)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(delta = runif(1,0.5,1), th2=runif(1),
                           th3 = runif(1), th1=runif(1,0,0.2),
                           sigma = runif(1))
    parameters <- c("th1", "th2","th3","th4","sigma")
    return(list(para=parameters, data=stan.dat, inits=inits,
                n.chains=chains))
}

test_stan2 <- list() # list for each test
for (i in 1:214){
    print(paste("Test:", i))
    input.to.stan <-
        stan.in2(data=NOAA5mean[NOAA5mean$Test==i,]) # evaluating by test
    fit2keep <- sampling(stan.fit1, data = input.to.stan$data,
                         init=input.to.stan$inits,
                         pars = input.to.stan$para,
                         iter=niters, thin=nthin,
                         chains=input.to.stan$n.chains)
    print(fit2keep)
    test_stan2[[i]] <- rstan::extract(fit2keep)
}
save(test_stan2, file="BC5.RData")
load("BC5.RData")
```

# Predictive uncretainty for IFE 5 and IFE 12
we extracted uncertainty even though they don't report. the estimate is highly variable. draw random samples from simulation.
```{r}
## Predictive uncertainty using inverse function 
##  The inverse function log(x) = log(th3) - log((th1-y)/(y-th4))/th2

predNOAA<- function(stanFit=test_stan1[[1]],
                       data=NOAA[NOAA$Test==1&is.na(NOAA$Concentration),],
                       meanABS=NOAA5mean$Absorbance[NOAA5mean$Concentration==0&NOAA5mean$Test==1],
                       mn=F){## mn=T -- 5 data points, mn=F -- 12 data points
    th1rv <- rvsims(stanFit$th1)
    th2rv <- rvsims(stanFit$th2)
    th3rv <- rvsims(stanFit$th3)
    th4rv <- rvsims(stanFit$th4)
    sigma <- rvsims(stanFit$sigma)
    if (!mn){
        data <- data
        m <- dim(data)[1]/2 # number of rows/2
        y <- data$Absorbance+rvnorm(mean=0, sd=sigma*sqrt((12+m-4)/(12-4)))
    } else{
        indtmp <- paste(data$SampleID, data$Dilution)
        tmp <- tapply(data$Absorbance, indtmp, mean)
        data <- data.frame(SampleID=names(tmp), Absorbance=tmp,
                           Dilution=tapply(data$Dilution, indtmp, mean))
        m <- dim(data)[1]/2 # number of rows/2
        y <- data$Absorbance/meanABS+rvnorm(mean=0,
                                            sd=sigma*sqrt((5+m-4)/(5-4)))
    }
    conc_pred <- exp(log(th3rv)-log((th1rv-y)/(y-th4rv))/th2rv)*data$Dilution
    temp <- summary(conc_pred)
    temp$SampleID <- data$SampleID
    temp$Dilution <- data$Dilution
    return(list(summ=temp, rvs=conc_pred))
}

### 12 point curve
pred_summ12 <- list()
pred_rvs12 <- list()
for (i in 1:214){
    temp <- predNOAA(stanFit=test_stan1[[i]],
                       data=NOAA[NOAA$Test==i&is.na(NOAA$Concentration),],
                       meanABS=NOAA5mean$Absorbance[NOAA5mean$Concentration==0&NOAA5mean$Test==i],
                       mn=F)
    pred_summ12[[i]] <- temp$summ
    pred_rvs12[[i]] <- temp$rvs
}

### 5 point curve 
pred_summ5 <- list()
pred_rvs5 <- list()
for (i in 1:214){
    temp <- predNOAA(stanFit=test_stan2[[i]],
                               data=NOAA[NOAA$Test==i&is.na(NOAA$Concentration),],
                               meanABS=NOAA5mean$Absorbance[NOAA5mean$Concentration==0&NOAA5mean$Test==i],
                               mn=T)
    pred_summ5[[i]] <- temp$summ
    pred_rvs5[[i]] <- temp$rvs
}

par(mar=c(3,3,1,0), mgp=c(1.25,0.125,0), las=1, tck=-0.01)
boxplot(log(pred_summ12[[5]]$sd), log(pred_summ5[[5]]$sd),
        names=c("$n=12$", "$n=5$"), ylab="log standard deviation")

save(pred_summ12, file="pred_summBC12.RData")
save(pred_rvs12, file="pred_rvsBC12.RData")
save(pred_summ5, file="pred_summBC5.RData")
save(pred_rvs5, file="pred_rvsBC5.RData")

load("pred_summBC12.RData")
load("pred_rvsBC12.RData")
load("pred_summBC5.RData")
load("pred_rvsBC5.RData")
```

#3) Bayesian (non-informative prior; Bayes)
Bayesian estimation with no hierarchical structure.
```{r}
### Default ELISA model with water samples
###    unknown concentrations estimated using MLE
### Stan Model (single test kit)
### The model: y=th4+(th1-th4)/[1+(x/th3)^(-th2)]

ELISA2 <- "
data {
  int<lower=0> N; // sample size (training data, non-zero)
  int<lower=0> M; //observed ODs (for estimating concentration)
  real y[N]; // response data
  real x[N]; // predictor data
  real y0[M]; // test data response
  int MM; //unique samples
  int ws[M]; //unique water samples
}
parameters {
  real<lower=0> delta;
  real<lower=0> th1;
  real<lower=0> th2;
  real<lower=0> th3;
  real<lower=0> sigma;
  vector[MM] logx0;
}
transformed parameters{
  real mu[N];
  real mu0[M];
  vector<lower=0>[MM] x0;
  real<lower=0> th4;
  th4 = th1+delta;
  x0 = exp(logx0);
  for (i in 1:N){
    mu[i] = th4 - delta/(1+(x[i]/th3)^(-th2));
  }
  for (i in 1:M){
    mu0[i] = th4 - delta/(1+((x0[ws[i]])/th3)^(-th2));
  }
}
model {
  logx0 ~ normal(0,2.5); // non-informative prior for all concentration value (no hierarchical structure). Centering logx0 at 0.
  delta~normal(1,1);
  th1~normal(0,1);
  th2~normal(0,5);
  th3~normal(0,5);
  target += normal_lpdf(y| mu, sigma);
  target += normal_lpdf(y0| mu0, sigma);
}
"
stan.fit2 <- stan_model(model_code=ELISA2)

stan.in3 <- function(data = NOAA, chains=nchains){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    y <- data$Absorbance[!temp]
    x <- data$Concentration[!temp]
    N <- n-sum(temp)
    M <- sum(temp)
    y0 <- data$Absorbance[temp]
    wsC<-ordered(data$SampleID[temp])
    ws <- as.numeric(wsC)
    stan.dat <- list(N=N, M=M, MM=max(ws), y=y, x=x,
                     y0=y0, ws=ws)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(delta = runif(1,0.5,1), th2=runif(1),
                           th3 = runif(1), th1=runif(1,0,0.1),
                           logx0 = rnorm(max(ws)),
                           sigma = runif(1))
    parameters <- c("th1","th2","th3","th4","sigma","x0")
    return(list(para=parameters, data=stan.dat, inits=inits,
              n.chains=chains, ControlSample=(1:max(ws))[levels(wsC)=="Control 1"]))
}

test_stanMLE1 <- list()
controlS <- numeric()
for (i in 1:214){
    print(paste("Test:", i))
    input.to.stan <- stan.in3(data=NOAA[NOAA$Test==i,])
    if(length(input.to.stan$ControlSample)==0) # added 8/2/22
    controlS[i]<-0
    else
    controlS[i] <- input.to.stan$ControlSample
}
for (i in 1:214){
    print(paste("Test:", i))
    input.to.stan <- stan.in3(data=NOAA[NOAA$Test==i,])
    fit2keep <- sampling(stan.fit2, data = input.to.stan$data,
                         init=input.to.stan$inits,
                         pars = input.to.stan$para,
                         iter=niters, thin=nthin,
                         chains=input.to.stan$n.chains)
    print(fit2keep)
    test_stanMLE1[[i]] <- rstan::extract(fit2keep)
    if(length(input.to.stan$ControlSample)==0) # added 8/2/22
    controlS[i]<-0
    else
    controlS[i] <- input.to.stan$ControlSample
}
save(test_stanMLE1, file="Bayes.RData")
load("Bayes.RData")
```

# 4) BHM within test (BHM 1)
hierarchical model estimating water sample. hierarchical structure imposed on log concentration.

We estimate water samples. We find the average of the water samples. We center them around the average (like the baseball example). Large concentrations must be diluted to fall within the curve
```{r}
### Default ELISA model with water samples
### unknown concentrations estimated using hierarchical model
### Stan Model (single test kit)
### The model: y=th4+(th1-th4)/[1+(x/th3)^(-th2)]

ELISA3 <- "
data {
  int<lower=0> N; // sample size (training data, non-zero)
  int<lower=0> M; //observed ODs (for estimating concentration)
  real y[N]; // response data
  real x[N]; // predictor data
  real y0[M]; // test data response
  int MM; //unique samples
  int ws[M]; //unique water samples
}
parameters {
  real<lower=0> delta;
  real<lower=0> th1;
  real<lower=0> th2;
  real<lower=0> th3;
  real<lower=0> sigma;
  vector[MM] logx0;
  real<lower=0> x0sig;
  real x0mu;
}
transformed parameters{
  vector[N] mu;
  vector[M] mu0;
  vector<lower=0>[MM] x0;
  real<lower=0> th4;
  x0 = exp(logx0);
  th4 = th1+delta;
  for (i in 1:N){
    mu[i] = th4 - delta/(1+(x[i]/th3)^(-th2));
  }
  for (i in 1:M){
    mu0[i] = th4 - delta/(1+((x0[ws[i]])/th3)^(-th2)); 
  }
}
model {
  x0mu ~ normal(0, 2.5); // non-informative prior for log conc mean
  x0sig ~ normal(0, 2.5); // non-informative prior for log conc variance
  logx0 ~ normal(x0mu, x0sig); // not centering log x0 at 1, instead center of all of the x0 means. 
  
  delta~normal(1,1);
  th1~normal(0,1); // th1 can't be too big so we constrain the prior a bit
  th2~normal(0,5);
  th3~normal(0,5);
  target += normal_lpdf(y| mu, sigma); // mu is provided by data
  target += normal_lpdf(y0| mu0, sigma); // mu is entered by parameter we need to estimate
}
"
stan.fit3 <- stan_model(model_code=ELISA3)

stan.in4 <- function(data = NOAA, chains=nchains){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    y <- data$Absorbance[!temp]
    x <- data$Concentration[!temp]
    N <- n-sum(temp)
    M <- sum(temp)
    y0 <- data$Absorbance[temp]
    wsC <- ordered(data$SampleID[temp])
    ws <- as.numeric(wsC)
    stan.dat <- list(N=N, M=M, MM=max(ws), y=y, x=x,
                     y0=y0, ws=ws)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(delta = runif(1, 0.5, 1), th2=runif(1, 0.4, 0.6),
                           th3 = runif(1, 0.3, 0.7), th1=runif(1, 0, 0.1),
                           logx0 = rnorm(max(ws)),
                           sigma = runif(1), x0mu=rnorm(1), x0sig=runif(1))
    parameters <- c("th1", "th2","th3","th4","sigma","x0", "x0mu","x0sig")
    return(list(para=parameters, data=stan.dat, inits=inits, n.chains=chains,
                ControlSample=(1:max(ws))[levels(wsC)=="Control 1"]))
}

test_stanBHM1 <- list()
for (i in 1:214){
    print(paste("Test:", i))
    input.to.stan <- stan.in4(data=NOAA[NOAA$Test==i,])
    fit2keep <- sampling(stan.fit3, data = input.to.stan$data,
                         init=input.to.stan$inits,
                         pars = input.to.stan$para,
                         iter=niters, thin=nthin,
                         chains=input.to.stan$n.chains)
    print(fit2keep)
    test_stanBHM1[[i]] <- rstan::extract(fit2keep)
}
save(test_stanBHM1, file="BHM1.RData")
load("BHM1.RData") 

print(test_stanBHM1[[1]])
plot(test_stanBHM1[[1]]$x0mu, test_stanBHM1[[1]]$x0sig) 
```

# load priors - established using the first 9 tests. File with this code is called 'ELISA_BHM_Prior'

```{r}
load("BHM_Prior.RData")
```
#5) BHM across test (sequential updating; BHM 2), not sharing residual variance across tests
hierarchical structure only across test curve coefficients
```{r}
### Sequential updating ELISA model with water samples
### unknown concentrations estimated using hierarchical model (across test only)
### The model: y=th4+(th1-th4)/[1+(x/th3)^(-th2)]

ELISA5 <- "
data {
  int<lower=0> N; // sample size (training data, non-zero)
  int<lower=0> M; //observed ODs (for estimating concentration)
  real y[N]; // response data
  real x[N]; // predictor data
  real y0[M]; // test data response
  int MM; //unique samples
  int ws[M]; //unique water samples
  real minABS; //minimum observed absorbance

  real deltamu0; //prior means
  real th1mu0;
  real th2mu0;
  real th3mu0;
  real<lower=0> deltan0; // prior lambda
  real<lower=0> th1n0;
  real<lower=0> th2n0;
  real<lower=0> th3n0;
  real<lower=0> deltalpha; // prior alpha
  real<lower=0> th1alpha;
  real<lower=0> th2alpha;
  real<lower=0> th3alpha;
  real<lower=0> deltabeta; // prior beta
  real<lower=0> th1beta;
  real<lower=0> th2beta;
  real<lower=0> th3beta;
  real<lower=0> aa; // prior a (sigma_y)
  real<lower=0> bb; // prior b
}
parameters {
  real<lower=0> delta;
  real<upper=minABS> logth1;
  real<lower=0> th2;
  real<lower=0> th3;
  real<lower=0> sig2;
  vector[MM] logx0;
  real mu_dlt;
  real<lower=0> sig2_dlt;
  real mu_th1;
  real<lower=0> sig2_th1;
  real mu_th2;
  real<lower=0> sig2_th2;
  real mu_th3;
  real<lower=0> sig2_th3;
}
transformed parameters{
  vector[N] mu;
  vector[M] mu0;
  real th1;
  real<lower=0> th4;
  vector<lower=0>[MM] x0;
  x0 = exp(logx0);
  th1 = exp(logth1);
  th4 = th1+delta;
  for (i in 1:N){
    mu[i] = th4 - delta/(1+(x[i]/th3)^(-th2));
  }
  for (i in 1:M){
    mu0[i] = th4 - delta/(1+(x0[ws[i]]/th3)^(-th2));
  }
}
model {
  logx0~normal(0, 2.5);

  sig2 ~ inv_gamma(aa, bb);
  sig2_dlt ~ inv_gamma(deltalpha, deltabeta);
  mu_dlt ~ normal(deltamu0, sqrt(sig2_dlt/deltan0));
  sig2_th1 ~ inv_gamma(th1alpha, th1beta);
  mu_th1 ~ normal(th1mu0, sqrt(sig2_th1/th1n0));
  sig2_th2 ~ inv_gamma(th2alpha, th2beta);
  mu_th2 ~ normal(th2mu0, sqrt(sig2_th2/th2n0));
  sig2_th3 ~ inv_gamma(th3alpha, th3beta);
  mu_th3 ~ normal(th3mu0, sqrt(sig2_th3/th3n0));

  delta~normal(mu_dlt,sqrt(sig2_dlt)); // the difference
  logth1~normal(mu_th1, sqrt(sig2_th1)); // th1 must be bigger than zero bc it is the minimum abs. That is why it is log.
  th2~normal(mu_th2, sqrt(sig2_th2)); // shape
  th3~normal(mu_th3, sqrt(sig2_th3)); // mid 

  target += normal_lpdf(y| mu, sqrt(sig2));
  target += normal_lpdf(y0| mu0, sqrt(sig2));
}
"

stan.fit6 <- stan_model(model_code=ELISA5)

stan.in7 <- function(data = NOAA[NOAA$Test==1,], chains=nchains,
                     Dstan, th1stan, th2stan, th3stan, sig2stan){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    tmp1 <- data[!temp,]
    tmp2 <- data[temp&data$Absorbance<100,]
    test1 <- as.numeric(ordered(tmp1$Test))
    y <- tmp1$Absorbance
    x <- tmp1$Concentration
    N <- dim(tmp1)[1]
    M <- dim(tmp2)[1]
    y0 <- tmp2$Absorbance
    wsC <- ordered(paste(tmp2$Test,tmp2$SampleID))
    wsLevels <- levels(wsC)
    tmpp<-str_detect(wsLevels, "Control")
    ws <- as.numeric(wsC)
    stan.dat <- list(N=N, M=M, MM=max(ws), y=y, x=x,
                     y0=y0, ws=ws, minABS=log(min(c(y0,y))),
                     deltamu0=Dstan$mu0, th1mu0=th1stan$mu0,
                     th2mu0=th2stan$mu0, th3mu0=th3stan$mu0,
                     deltan0=Dstan$lambda, th1n0=th1stan$lambda,
                     th2n0=th2stan$lambda, th3n0=th3stan$lambda,
                     deltalpha=Dstan$alpha, th1alpha=th1stan$alpha,
                     th2alpha=th2stan$alpha, th3alpha=th3stan$alpha,
                     deltabeta=Dstan$beta, th1beta=th1stan$beta,
                     th2beta=th2stan$beta, th3beta=th3stan$beta,
                     aa=0.001, bb=0.001) # not sharing residual variance 
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(zdelta=runif(1,0.5,1), zth1=runif(1,0,0.1),
                           zth2=runif(1,0.4,0.6), zth3=runif(1,0.3,0.7),
                           sig2=runif(1), logx0=rnorm(max(ws)),
                           mu_dlt=runif(1), sig2_dlt=runif(1),
                           mu_th1=runif(1), sig2_th1=runif(1),
                           mu_th2=runif(1), sig2_th2=runif(1),
                           mu_th3=runif(1), sig2_th3=runif(1))

    parameters <- c("th1", "th2","th3","th4","sig2","x0",
                    "mu_dlt", "sig2_dlt", "mu_th1",
                    "sig2_th1", "mu_th2", "sig2_th2", "mu_th3", "sig2_th3")
    return(list(para=parameters, data=stan.dat, inits=inits, n.chains=chains,
                ControlSample=(1:max(ws))[tmpp], samplenames=wsLevels, wsID=1:max(ws)))
}

prior_pars_NIG <- function(mus, sigs2){
  Ex <- mean(mus)
  Vx <- sd(mus)^2
  Esig2 <- mean(sigs2)
  Vsig2 <- sd(sigs2)^2
  return(list(mu0=Ex, beta=Esig2*(1+Esig2^2/Vsig2),
              alpha=2+Esig2^2/Vsig2, lambda=Esig2/Vx))
}

prior_pars_IG <- function(sig2){
    Esig2 <- mean(sig2)
    Vsig2 <- sd(sig2)^2
    return(list(alpha=2+Esig2^2/Vsig2, beta=Esig2*(1+Esig2^2/Vsig2)))
}

sig2_stan <- prior_pars_IG(test_stanBHM2$sigma^2)
D_stan <- prior_pars_NIG(test_stanBHM2$mu_dlt, test_stanBHM2$sig_dlt^2)
th1_stan <- prior_pars_NIG(test_stanBHM2$mu_th1, test_stanBHM2$sig_th1^2)
th2_stan <- prior_pars_NIG(test_stanBHM2$mu_th2, test_stanBHM2$sig_th2^2)
th3_stan <- prior_pars_NIG(test_stanBHM2$mu_th3, test_stanBHM2$sig_th3^2)

seqBHM2 <- list()
seqBHM2_all<-list()
BHM2_IDs<-list()
for (i in 10:214){
    print(paste("Test: ", i))
          input.to.stan <- stan.in7(data=NOAA[NOAA$Test==i,],
                              Dstan=D_stan, th1stan=th1_stan,
                              th2stan=th2_stan, th3stan=th3_stan,
                              sig2stan=sig2_stan)

    fit2keep <- sampling(stan.fit6, data = input.to.stan$data,
                         init=input.to.stan$inits,
                         pars = input.to.stan$para,
                         iter=niters, thin=nthin,
                         chains=input.to.stan$n.chains)
    print(fit2keep)
    seqBHM2_all[[i]]<-fit2keep
    seqBHM2[[i]]<- rstan::extract(fit2keep)
    sig2_stan <- prior_pars_IG(seqBHM2[[i]]$sig2)
    D_stan <- prior_pars_NIG(seqBHM2[[i]]$mu_dlt, seqBHM2[[i]]$sig2_dlt)
    th1_stan <- prior_pars_NIG(seqBHM2[[i]]$mu_th1, seqBHM2[[i]]$sig2_th1)
    th2_stan <- prior_pars_NIG(seqBHM2[[i]]$mu_th2, seqBHM2[[i]]$sig2_th2)
    th3_stan <- prior_pars_NIG(seqBHM2[[i]]$mu_th3, seqBHM2[[i]]$sig2_th3)
    
    #save ID names
    wsEstimate<-cbind(as.data.frame(input.to.stan$wsID, input.to.stan$samplenames))
    xOD2<-rvsims(as.matrix(as.data.frame(extract(fit2keep, pars="x0"))))
    wsEstimate<-cbind(wsEstimate, as.data.frame(summary(xOD2))) 
    wsEstimate<-rownames_to_column(wsEstimate, "SampleID") 
    BHM2_IDs[[i]]<-wsEstimate
}

save(seqBHM2, file="BHM2.RData")
#save(BHM2_IDs, file="BHM2_IDs.RData")
#save(seqBHM2_all, file="BHM2_seqBHMall.RData")
load("BHM2.RData")
#load("BHM2_IDs.RData")
#load("BHM2_seqBHMall.RData")

```

#6) BHM within and across test (sequential updating; BHM 3), not sharing residual variance across tests
```{r}
### Sequential updating ELISA model with water samples
###    unknown concentrations estimated using hierarchical model
### Stan Model (single multiple test kit)
### The model: y=th4+(th1-th4)/[1+(x/th3)^(-th2)]

ELISA6 <- "
data {
  int<lower=0> N; // sample size (training data, non-zero)
  int<lower=0> M; //observed ODs (for estimating concentration)
  real y[N]; // response data
  real x[N]; // predictor data
  real y0[M]; // test data response
  int MM; //unique samples
  int ws[M]; //unique water samples
  real minABS; //minimum observed absorbance

  real deltamu0; //prior means
  real th1mu0;
  real th2mu0;
  real th3mu0;
  real<lower=0> deltan0; // prior lambda
  real<lower=0> th1n0;
  real<lower=0> th2n0;
  real<lower=0> th3n0;
  real<lower=0> deltalpha; // prior alpha
  real<lower=0> th1alpha;
  real<lower=0> th2alpha;
  real<lower=0> th3alpha;
  real<lower=0> deltabeta; // prior beta
  real<lower=0> th1beta;
  real<lower=0> th2beta;
  real<lower=0> th3beta;
  real<lower=0> aa; // prior a (sigma_y)
  real<lower=0> bb; // prior b
}
parameters {
  real<lower=0> delta;
  real<upper=minABS> logth1;
  real<lower=0> th2;
  real<lower=0> th3;
  real<lower=0> sig2;
  vector[MM] zlogx0;
  real<lower=0> x0sig;
  real x0mu;
  real mu_dlt;
  real<lower=0> sig2_dlt;
  real mu_th1;
  real<lower=0> sig2_th1;
  real mu_th2;
  real<lower=0> sig2_th2;
  real mu_th3;
  real<lower=0> sig2_th3;
}
transformed parameters{
  vector[N] mu;
  vector[M] mu0;
  real th1;
  real<lower=0> th4;
  vector<lower=0>[MM] x0;
  vector[MM] logx0;
  logx0 = x0mu + x0sig*zlogx0;
  x0 = exp(logx0);
  th1=exp(logth1);
  th4 = th1+delta;
  for (i in 1:N){
    mu[i] = th4 - delta/(1+(x[i]/th3)^(-th2));
  }
  for (i in 1:M){
    mu0[i] = th4 - delta/(1+(x0[ws[i]]/th3)^(-th2));
  }
}
model {
  x0mu ~ normal(0, 2.5);
  x0sig ~ normal(0, 2.5);
  zlogx0 ~ std_normal(); // z computational trick

  sig2 ~ inv_gamma(aa, bb);
  sig2_dlt ~ inv_gamma(deltalpha, deltabeta);
  mu_dlt ~ normal(deltamu0, sqrt(sig2_dlt/deltan0));
  sig2_th1 ~ inv_gamma(th1alpha, th1beta);
  mu_th1 ~ normal(th1mu0, sqrt(sig2_th1/th1n0));
  sig2_th2 ~ inv_gamma(th2alpha, th2beta);
  mu_th2 ~ normal(th2mu0, sqrt(sig2_th2/th2n0));
  sig2_th3 ~ inv_gamma(th3alpha, th3beta);
  mu_th3 ~ normal(th3mu0, sqrt(sig2_th3/th3n0));

  delta~normal(mu_dlt,sqrt(sig2_dlt));
  logth1~normal(mu_th1, sqrt(sig2_th1)); // mu_th1 is for logth1--opens up the lower end, doesn't constrain high
  th2~normal(mu_th2, sqrt(sig2_th2)); // mu_th2 is for th2 itself
  th3~normal(mu_th3, sqrt(sig2_th3));

  target += normal_lpdf(y| mu, sqrt(sig2));
  target += normal_lpdf(y0| mu0, sqrt(sig2));
}
"


stan.fit5 <- stan_model(model_code=ELISA6)

stan.in6 <- function(data = NOAA[NOAA$Test==10,], chains=nchains,
                     Dstan, th1stan, th2stan, th3stan, sig2stan){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    tmp1 <- data[!temp,]
    tmp2 <- data[temp,]
    test1 <- as.numeric(ordered(tmp1$Test))
    y <- tmp1$Absorbance
    x <- tmp1$Concentration
    N <- dim(tmp1)[1]
    M <- dim(tmp2)[1]
    y0 <- tmp2$Absorbance
    wsC <- ordered(paste(tmp2$Test,tmp2$SampleID))
    wsLevels <- levels(wsC)
    tmpp<-str_detect(wsLevels, "Control")
    ws <- as.numeric(wsC)
    stan.dat <- list(N=N, M=M, MM=max(ws), y=y, x=x,
                     y0=y0, ws=ws, minABS=log(min(c(y0,y))),
                     deltamu0=Dstan$mu0, th1mu0=th1stan$mu0,
                     th2mu0=th2stan$mu0, th3mu0=th3stan$mu0,
                     deltan0=Dstan$lambda, th1n0=th1stan$lambda,
                     th2n0=th2stan$lambda, th3n0=th3stan$lambda,
                     deltalpha=Dstan$alpha, th1alpha=th1stan$alpha,
                     th2alpha=th2stan$alpha, th3alpha=th3stan$alpha,
                     deltabeta=Dstan$beta, th1beta=th1stan$beta,
                     th2beta=th2stan$beta, th3beta=th3stan$beta,
                     aa=0.001, bb=0.001) # not sharing residual variance
                     
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(zdelta=runif(1,0.5,1), zth1=runif(1,0,0.1),
                           zth2=runif(1,0.4,0.6), zth3=runif(1,0.3,0.7),
                           sig2=runif(1), zlogx0=rnorm(max(ws)),
                           x0sig=runif(1), x0mu=runif(1),
                           mu_dlt=runif(1), sig2_dlt=runif(1),
                           mu_th1=runif(1), sig2_th1=runif(1),
                           mu_th2=runif(1), sig2_th2=runif(1),
                           mu_th3=runif(1), sig2_th3=runif(1))

    parameters <- c("th1", "th2","th3","th4","sig2","x0",
                    "x0mu","x0sig", "mu_dlt", "sig2_dlt", "mu_th1",
                    "sig2_th1", "mu_th2", "sig2_th2", "mu_th3", "sig2_th3")
    return(list(para=parameters, data=stan.dat, inits=inits, n.chains=chains,
                ControlSample=(1:max(ws))[tmpp], samplenames=wsLevels, wsID=1:max(ws)))
}

prior_pars_NIG <- function(mus, sigs2){
  Ex <- mean(mus)
  Vx <- sd(mus)^2
  Esig2 <- mean(sigs2)
  Vsig2 <- sd(sigs2)^2
  return(list(mu0=Ex, beta=Esig2*(1+Esig2^2/Vsig2),
              alpha=2+Esig2^2/Vsig2, lambda=Esig2/Vx))
}

prior_pars_IG <- function(sig2){
    Esig2 <- mean(sig2)
    Vsig2 <- sd(sig2)^2
    return(list(alpha=2+Esig2^2/Vsig2, beta=Esig2*(1+Esig2^2/Vsig2)))
}

sig2_stan <- prior_pars_IG(test_stanBHM2$sigma^2)
D_stan <- prior_pars_NIG(test_stanBHM2$mu_dlt, test_stanBHM2$sig_dlt^2)
th1_stan <- prior_pars_NIG(test_stanBHM2$mu_th1, test_stanBHM2$sig_th1^2)
th2_stan <- prior_pars_NIG(test_stanBHM2$mu_th2, test_stanBHM2$sig_th2^2)
th3_stan <- prior_pars_NIG(test_stanBHM2$mu_th3, test_stanBHM2$sig_th3^2)

seqBHM3 <- list()
seqBHM3all<-list()
BHM3_IDs<-list()
for (i in 10:214){ # start from test 10 b/c 1-9 is used to create initial priors
    print(paste("Test: ", i))
          input.to.stan <- stan.in6(data=NOAA[NOAA$Test==i,],
                              Dstan=D_stan, th1stan=th1_stan,
                              th2stan=th2_stan, th3stan=th3_stan,
                              sig2stan=sig2_stan)

    fit2keep <- sampling(stan.fit5, data = input.to.stan$data,
                         init=input.to.stan$inits,
                         pars = input.to.stan$para,
                         iter=niters, thin=nthin,
                         chains=input.to.stan$n.chains)
    print(fit2keep)
    #j<-i-9
    seqBHM3all[[i]]<-fit2keep
    seqBHM3[[i]]<- rstan::extract(fit2keep)
    sig2_stan <- prior_pars_IG(seqBHM3[[i]]$sig2)
    D_stan <- prior_pars_NIG(seqBHM3[[i]]$mu_dlt, seqBHM3[[i]]$sig2_dlt)
    th1_stan <- prior_pars_NIG(seqBHM3[[i]]$mu_th1, seqBHM3[[i]]$sig2_th1)
    th2_stan <- prior_pars_NIG(seqBHM3[[i]]$mu_th2, seqBHM3[[i]]$sig2_th2)
    th3_stan <- prior_pars_NIG(seqBHM3[[i]]$mu_th3, seqBHM3[[i]]$sig2_th3)
    
    #save ID names
    wsEstimate<-cbind(as.data.frame(input.to.stan$wsID, input.to.stan$samplenames))
    xOD2<-rvsims(as.matrix(as.data.frame(extract(fit2keep, pars="x0"))))
    wsEstimate<-cbind(wsEstimate, as.data.frame(summary(xOD2))) 
    wsEstimate<-rownames_to_column(wsEstimate, "SampleID") 
    BHM3_IDs[[i]]<-wsEstimate
}

save(seqBHM3, file="BHM3.RData")
#save(BHM3_IDs, file="BHM3_IDs.RData")
#save(seqBHM3all, file="BHM3_seqBHMall.RData")
load("BHM3.RData")
#load("BHM3_IDs.RData")
#load("BHM3_seqBHMall.RData")

```

# Model comparison plots
Load model results
```{r}
# model 1
load("BC5.RData") 
load("pred_summBC5.RData")
load("pred_rvsBC5.RData")
# model 2
load("BC12.RData") 
load("pred_summBC12.RData")
load("pred_rvsBC12.RData")
# model 3
load("Bayes.RData")
# model 4
load("BHM1.RData")
# model 5
load("BHM2.RData")
# model 6
load("BHM3.RData")
```
load in 'controlS' which are needed for the plots. This code comes from the Bayes model.
```{r}
stan.in3 <- function(data = NOAA, chains=nchains){
    n <- dim(data)[1]
    temp  <- is.na(data$Concentration)
    y <- data$Absorbance[!temp]
    x <- data$Concentration[!temp]
    N <- n-sum(temp)
    M <- sum(temp)
    y0 <- data$Absorbance[temp]
    wsC<-ordered(data$SampleID[temp])
    ws <- as.numeric(wsC)
    stan.dat <- list(N=N, M=M, MM=max(ws), y=y, x=x,
                     y0=y0, ws=ws)
    inits <- list()
    for (i in 1:chains)
        inits[[i]] <- list(delta = runif(1,0.5,1), th2=runif(1),
                           th3 = runif(1), th1=runif(1,0,0.1),
                           logx0 = rnorm(max(ws)),
                           sigma = runif(1))
    parameters <- c("th1","th2","th3","th4","sigma","x0")
    return(list(para=parameters, data=stan.dat, inits=inits,
              n.chains=chains, ControlSample=(1:max(ws))[levels(wsC)=="Control 1"]))
}

controlS <- numeric()
for (i in 1:214){
    #print(paste("Test:", i))
    input.to.stan <- stan.in3(data=NOAA[NOAA$Test==i,])
    if(length(input.to.stan$ControlSample)==0) # added 8/2/22
    controlS[i]<-0
    else
    controlS[i] <- input.to.stan$ControlSample
}
```
Compare the control estimates for each model bc controls have a known concentration of 0.75.
```{r}
## comparing control sample fits

cntrlID <- (1:length(pred_summ12[[1]]))[substring(pred_summ12[[i]]$SampleID, 1, 9)=="Control 1"]
CntrlMLE12 <- pred_summ12[[1]][cntrlID,]
CntrlMLErvs12 <- pred_rvs12[[1]][cntrlID]

for (i in 2:214){
    cntrlID <- (1:length(pred_summ12[[i]]))[substring(pred_summ12[[i]]$SampleID, 1, 9)=="Control 1"]
    CntrlMLE12 <- rbind(CntrlMLE12[,1:9], pred_summ12[[i]][cntrlID,1:9])
    CntrlMLErvs12 <- c(CntrlMLErvs12, pred_rvs12[[i]][cntrlID])
}

cntrlID <- (1:dim(pred_summ5[[1]])[1])[substring(pred_summ5[[1]]$SampleID, 1, 9)=="Control 1"]
CntrlMLE5 <- pred_summ5[[1]][cntrlID,]
CntrlMLErvs5 <- pred_rvs5[[1]][cntrlID]
for (i in 2:214){
    cntrlID <- (1:dim(pred_summ5[[i]])[1])[substring(pred_summ5[[i]]$SampleID, 1, 9)=="Control 1"]
    CntrlMLE5 <- rbind(CntrlMLE5[,1:9], pred_summ5[[i]][cntrlID,1:9])
    CntrlMLErvs5 <- c(CntrlMLErvs5, pred_rvs5[[i]][cntrlID])
}

##Bayesian models
# MLE
CntrlStan1 <- rvsims(test_stanMLE1[[1]]$x0)[controlS[1]]
for (i in 2:214)
CntrlStan1 <- c(CntrlStan1, rvsims(test_stanMLE1[[i]]$x0)[controlS[i]])

# within-test only
CntrlStan3 <- rvsims(test_stanBHM1[[1]]$x0)[controlS[1]]
for (i in 2:214)
CntrlStan3 <- c(CntrlStan3, rvsims(test_stanBHM1[[i]]$x0)[controlS[i]])

# across test only sequential BHM (starts from test 10)
CntrlStanSeq2 <- rvsims(seqBHM2[[10]]$x0)[controlS[10]]
for (i in 11:214)
CntrlStanSeq2 <- c(CntrlStanSeq2, rvsims(seqBHM2[[i]]$x0)[controlS[i]])

# within and across test sequential BHM (starts from test 10)
CntrlStanSeq3 <- rvsims(seqBHM3[[10]]$x0)[controlS[10]]
for (i in 11:214)
CntrlStanSeq3 <- c(CntrlStanSeq3, rvsims(seqBHM3[[i]]$x0)[controlS[i]])
```
Comparing accuracy (0.75)
```{r}
accuracy <- c(median(abs(CntrlStanSeq3-0.75)),
          median(abs(CntrlStanSeq2-0.75)),
          median(abs(CntrlStan3-0.75)),
          median(abs(CntrlStan1-0.75)),
          median(abs(CntrlMLErvs12-0.75),na.rm=T),
          median(abs(CntrlMLErvs5-0.75),na.rm=T))
names(accuracy) <- c("BHM 3","BHM 2" ,"BHM 1",
                 "Bayes","IFE 12", "IFE 5")

par(mar=c(3, 3, 2, 0.125), mgp=c(1.25,0.25,0), tck=-0.01)
mlplot(accuracy, xlab="Accuracy", xlim=c(0.08,0.35), main="")

```

save accuracy comparison plot as pdf
```{r}
pdf(file = "6model_compare.pdf",   # The directory you want to save the file in
    width = 10, # The width of the plot in inches
    height = 7) # The height of the plot in inches
accuracy <- c(median(abs(CntrlStanSeq3-0.75)),
          median(abs(CntrlStanSeq2-0.75)),
          median(abs(CntrlStan3-0.75)),
          median(abs(CntrlStan1-0.75)),
          median(abs(CntrlMLErvs12-0.75),na.rm=T),
          median(abs(CntrlMLErvs5-0.75),na.rm=T))
names(accuracy) <- c("BHM 3","BHM 2" ,"BHM 1",
                 "Bayes","IFE 12", "IFE 5")

par(mar=c(3, 3, 2, 0.125), mgp=c(1.25,0.25,0), tck=-0.01)
mlplot(accuracy, xlab="Accuracy", xlim=c(0.08, 0.35), main="")
dev.off()
```


